---
title: "Historical First and Last User Selected Temperature"
author: "Wendy Horvat"
output: html_document
runtime: shiny
---

```{r setup, include=FALSE}
library(shiny)
library(tidyverse)
library(lubridate)
library(ggtext)
```

# Historical Temperature Threshold Explorer

This interactive document explores historical temperature threshold events in Portland, Oregon.

Use the slider below to choose the minimum temperature threshold. The graph will update automatically to show:

- Gray points: every day with a minimum temperature at or below the selected threshold
- Purple points: the first occurrence after July each year
- Blue points: the last occurrence before July each year

---

# Load the data

```{r}
csv_data <- read_csv(
  "data/portland_1871_11_2022-12_temp_prec.csv",
  show_col_types = FALSE
)

# Make sure DATE is a Date object
csv_data <- csv_data %>%
  mutate(DATE = as.Date(DATE))
```

---

# Select a Temperature Threshold

```{r}
sliderInput(
  inputId = "temperature_threshold",
  label = "Temperature Threshold (°F)",
  min = 20,
  max = 40,
  value = 32,
  step = 1
)
```

---

# Interactive Plot

```{r, fig.height=7}

renderPlot({

  temperature_threshold <- input$temperature_threshold

  processed_data <- csv_data %>%
    mutate(
      year = year(DATE),
      month = month(DATE),
      day_of_year = yday(DATE)
    )

  threshold_event_days <- processed_data %>%
    filter(TMIN <= temperature_threshold)

  first_threshold_event <- processed_data %>%
    group_by(year) %>%
    filter(
      TMIN <= temperature_threshold,
      month > 7
    ) %>%
    summarise(
      event_date = min(DATE),
      .groups = "drop"
    ) %>%
    mutate(
      day_of_year = yday(event_date)
    )

  last_threshold_event <- processed_data %>%
    group_by(year) %>%
    filter(
      TMIN <= temperature_threshold,
      month < 7
    ) %>%
    summarise(
      event_date = max(DATE),
      .groups = "drop"
    ) %>%
    mutate(
      day_of_year = yday(event_date)
    )

  ggplot() +

    geom_point(
      data = threshold_event_days,
      aes(
        x = as.Date(day_of_year, origin = "2020-01-01"),
        y = year
      ),
      color = "gray",
      size = 0.8
    ) +

    geom_point(
      data = first_threshold_event,
      aes(
        x = as.Date(day_of_year, origin = "2020-01-01"),
        y = year
      ),
      fill = "purple",
      color = "black",
      shape = 21,
      size = 3
    ) +

    geom_point(
      data = last_threshold_event,
      aes(
        x = as.Date(day_of_year, origin = "2020-01-01"),
        y = year
      ),
      fill = "blue",
      color = "black",
      shape = 21,
      size = 3
    ) +

    scale_y_reverse(
      breaks = seq(1870, 2020, 10),
      minor_breaks = seq(1870, 2020, 2)
    ) +

    scale_x_date(
      date_labels = "%b",
      breaks = "month"
    ) +

    theme_minimal() +

    labs(
      title = paste(
        "Annual First and Last Temperature Events at",
        temperature_threshold,
        "°F"
      ),
      subtitle = "Portland, Oregon",
      x = "",
      y = ""
    )

})
```

---

# About this Analysis

This visualization updates automatically whenever the temperature threshold changes.

The calculations are repeated each time the slider moves:

1. Identify every day with a minimum temperature at or below the selected threshold.
2. Find the final threshold event before July.
3. Find the first threshold event after July.
4. Plot all events on a yearly timeline.

Experiment with different thresholds to see how the timing of temperature events changes over the historical record.